---
title: "msImpute: Imputation of peptide intensity by low-rank approximation"
author: 
- name: Soroor Hediyeh-zadeh
  affiliation: &id The Walter and Eliza Hall Institute of Medical Research
- name: Andrew I. Webb
  affiliation: *id
- name: Melissa J. Davis
  affiliation: *id
package: msImpute
abstract: |
  MsImpute is a package for imputation of peptide intensity in proteomics experiments. It additionally contains tools for MAR/MNAR diagnosis and assessment of distortions to the probability distribution of the data post imputation. This document covers case studies featuring data obtained by various Mass Spectrometry (MS) acquisition modes to demonstrate applications of tools implemented in the package. The datasets selected here represent data with MAR and left-censored MNAR missingness patterns. We demonstrate msImpute is able to correctly identify these patterns and inform user's decisions in downstream analyses.
output: 
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
vignette: >
  %\VignetteIndexEntry{msImpute: proteomics missing values imputation and diagnosis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---




# Installation


```{r eval=FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("msImpute")
```


# Quick Start

The package consists of the following main functions:

- `selectFeatures`: identifies informative peptides that can be used to examine MAR/MNAR missingness in the data.

- `msImpute`: Main function that imputes missing values by learning a low-rank approximation of the data.

- `findVariableFeatures`: finds peptide with high biological variance. We use this in `computeStructuralMetrics`

- `computeStructuralMetrics`: returns a number of metrics that measure distortions into the data after imputation.

- `plotCV2`: Plots the square of coefficient of variation versus average log-expression i.e. mean-$CV^2$ plot


These functions overall are designed to inform user's decision in choosing a proper imputation strategy. For a more detailed workflow, please see [User's Manual](https://github.com/soroorh/proteomicscasestudies/blob/master/msImputeUsersGuide.pdf).


# TIMS Case Study: Blood plasma 

The aim is to assess the missing patterns in ion mobility data by Prianichnikov et al. (2020), available from PXD014777. The `evidence` table of MaxQuant output was processed as described below. Rows are Modified Peptide IDs. Charge state variations are treated as distinct peptide species. For peptides with multiple identification types, the intensity is considered to be the median of reported intensity values. Reverse complements and contaminant peptides are discarded. Peptides with more than 4 observed intensity values are retained. 

The data was acquired in two batches (over two days). We are interested to know if missing values are evenly distributed across batches, or there is a batch-specific dropout trend. The runs are also labeled by S1, S2 and S4 (source unknown). The aim is to use this information to work out if missing values occur due to technical or biological effects.  


```{r setup, message=FALSE}
library(reticulate)
library(msImpute)
library(limma)
library(imputeLCMD)
library(ComplexHeatmap)
```

## Data processing

The following procedures were applied to process the data, which we later load from the package data.

### Filter by detection


```{r}
data(pxd014777)
y <- pxd014777
```


Zero values that will be converted to Inf/-Inf after log- transformation. Check if there are valid values in the data before log transformation
```{r}
table(is.infinite(data.matrix(log2(y))))
```

There are zero values that will be converted to Inf/-Inf after log- transformation. Add a small offset to avoid infinite values:
```{r}
y <- log2(y+0.25)
```

### Normalization
```{r}
# quantile normalisation
y <- normalizeBetweenArrays(y, method = "quantile")
```

## Determine missing values pattern

Determine dominant patterns of missing values by investigating the distribution of missing values. Peptides that are missing in at least one experimental group (here batch), and therefore exhibit structured missing patterns can be identified by the EBM metric implemented in `selectFeatures`. We then make a heatmap of their dropout pattern.
```{r}

batch <- as.factor(gsub("(2018.*)_RF.*","\\1", colnames(y)))
experiment <- as.factor(gsub(".*(S[1-9]).*","\\1", colnames(y)))


hdp <- selectFeatures(y, method = "ebm", group = batch)


# peptides missing in one or more experimental group will have a NaN EBM, which is a measure of entropy of 
# distribution of observed values
table(is.nan(hdp$EBM))

# construct matrix M to capture missing entries
M <- ifelse(is.na(y),1,0)
M <- M[hdp$msImpute_feature,]

# plot a heatmap of missingness patterns for the selected peptides



```

```{r fig.cap="Heatmap of missing value patterns for peptides selected as informative peptides", fig.align="center"}
ha_column <- HeatmapAnnotation(batch = batch,
                               experiment = experiment,
                               col = list(batch = c('20181023' = "#B24745FF",
                                                    '20181024'= "#00A1D5FF"),
                                          experiment=c("S1"="#DF8F44FF",
                                                       "S2"="#374E55FF",
                                                       "S4"="#79AF97FF")))

hm <- Heatmap(M,
column_title = "dropout pattern, columns ordered by dropout similarity",
              name = "Intensity",
              col = c("#8FBC8F", "#FFEFDB"),
              show_row_names = FALSE,
              show_column_names = FALSE,
              cluster_rows = TRUE,
              cluster_columns = TRUE,
              show_column_dend = FALSE,
              show_row_dend = FALSE,
              top_annotation = ha_column,
              row_names_gp =  gpar(fontsize = 7),
              column_names_gp = gpar(fontsize = 8),
              heatmap_legend_param = list(#direction = "horizontal",
              heatmap_legend_side = "bottom",
              labels = c("observed","missing"),
              legend_width = unit(6, "cm")),
         )
hm <- draw(hm, heatmap_legend_side = "left")
```
The larger the EBM, the more scattered the missing values will be. If missing values are scattered across samples, their value can be estimated from the neighborhood, hence missing type is likely MNAR. If however, peptides are missing completely in one experimental condition, or they have much more concentrated (or dense) distributions, their EBM value will be lower. A `NaN` EBM suggests peptide is missing in at least one experimental group, defined by the `group` argument. Since there are 103 such peptides with `EBM=NaN`, this data
has peptides that are missing not at random i.e. the missingness is batch-specific. Given that this is a technical dataset, MNAR missing here can not be biological, and reflects batch-to-batch variations, such as differences in limit of detection etc.
`selectFeatures` just enables to detect any peptides that appear to exhibit structured missing, and hence might be left-censored.
you can also set `method="hvp"` which will select top `n_features` peptides with high dropout rate, defined as proportion of samples where a given peptide is missing, that are also highly expressed.


# DDA Case Study: Extracellular vesicles isolated from inflammatory bowel disease patients and controls

The study aims to characterize the proteomic profile of extracellular vesicles isolated from the descending colon of pediatric patients with inflammatory bowel disease and control participants. The following analysis is based on the `peptide` table from MaxQuant output, available from PXD007959. Rows are Modified Peptide IDs. Charge state variations are treated as distinct peptide species. Reverse complements and contaminant peptides are discarded. Peptides with more than 4 observed intensity values are retained. Additionally, qualified peptides are required to map uniquely to proteins. Two of the samples with missing group annotation were excluded.

## Filter by detection


The sample descriptions can be accessed via `pxd007959$samples`. Intensity values are stored in `pxd007959$y`.
```{r}
data(pxd007959)

sample_annot <- pxd007959$samples
y <- pxd007959$y
y <- log2(y)
```

## Normalization
We apply `cyclic loess` normalisation from `limma` to normalise log-intensities. We have justified use of `cyclic loess` method in depth in the user's guide.
```{r}
y <- normalizeBetweenArrays(y, method = "cyclicloess")
```


## Determine missing values pattern

```{r fig.align="center"}
# determine missing values pattern
group <- sample_annot$group
hdp <- selectFeatures(y, method="ebm", group = group)
```



```{r fig.cap="Dropout pattern of informative peptides", fig.align="center"}
# construct matrix M to capture missing entries
M <- ifelse(is.na(y),1,0)
M <- M[hdp$msImpute_feature,]



# plot a heatmap of missingness patterns for the selected peptides
ha_column <- HeatmapAnnotation(group = as.factor(sample_annot$group),
                               col=list(group=c('Control' = "#E64B35FF",
                                                'Mild' = "#3C5488FF",
                                                'Moderate' = "#00A087FF",
                                                'Severe'="#F39B7FFF")))

hm <- Heatmap(M,
column_title = "dropout pattern, columns ordered by dropout similarity",
              name = "Intensity",
              col = c("#8FBC8F", "#FFEFDB"),
              show_row_names = FALSE,
              show_column_names = FALSE,
              cluster_rows = TRUE,
              cluster_columns = TRUE,
              show_column_dend = FALSE,
              show_row_dend = FALSE,
              top_annotation = ha_column,
              row_names_gp =  gpar(fontsize = 7),
              column_names_gp = gpar(fontsize = 8),
              heatmap_legend_param = list(#direction = "horizontal",
              heatmap_legend_side = "bottom",
              labels = c("observed","missing"),
              legend_width = unit(6, "cm")),
         )
hm <- draw(hm, heatmap_legend_side = "left")
```
As it can be seen, samples from the control group cluster together. There is a structured, block-wise pattern of missing values in the 'Control' and 'Severe' groups. This suggests that missing in not at random. This is an example of **MNAR** dataset. Given this knowledge, we impute using `QRILC` and `msImpute`, setting method to `v2-mnar`. We then compare these methods by preservation of local (within experimental group) and global (between experimental group) similarities. Note that low-rank approximation generally works for data of MAR types. However, the algorithm implemented in `v2-mnar` makes it applicable to MNAR data. To make low-rank models applicable to 
MNAR data, we need to use it in a supervised mode, hence why we need to provide information about groups or biological/experimental
condition of each sample.

## Imputation
```{r}
# imputation

y_qrilc <- impute.QRILC(y)[[1]]

y_msImpute <- msImpute(y, method = "v2-mnar", group = group)



group <- as.factor(sample_annot$group)

```

## Assessment of preservation of local and global structures 

If you've installed python, and have set up a python environment in your session, you can run this section to compute the GW distance. Please see the user's guide for setup instructions. Note that you can still run `computeStructuralMetrics` by setting `y=NULL`, if there are no python environments setup.

**Withinness, betweenness and Gromov-Wasserstein (GW) distance**

`computeStructuralMerics` returns three metrics that can be used to compare various imputation procedures:

- `withinness` is the sum of the squared distances between samples from the same experimental group (e.g. control, treatment, Het, WT). More specifically the similarity of the samples is measured by the distance of the  (expression profile of the) sample from group centroid. This is a measure of preservation of local structures.

- `betweenness` is the sum of the squared distances between the experimental groups, more specifically the distance between group centroids. This is a measure of preservation of global structures.

- `gw_dist` is the Gromov-Wasserstein distance computed between Principal Components of imputed and source data. It is a measure of how well the structures are overall preserved over all principal axis of variation in the data. Hence, it captures preservation of both local and global structures. PCs of the source data are computed using highly variable peptides (i.e. peptides with high biological variance).

An ideal imputation method results in smaller `withinness`, larger `withinness` and smaller `gw_dist` among other imputation methods.

```{r eval=FALSE}
virtualenv_create('msImpute-reticulate')
virtualenv_install("msImpute-reticulate","scipy")
virtualenv_install("msImpute-reticulate","cython")
virtualenv_install("msImpute-reticulate","POT")

use_virtualenv("msImpute-reticulate")

top.hvp <- findVariableFeatures(y)

computeStructuralMetrics(y_msImpute, group, y[rownames(top.hvp)[1:50],], k = 16)
```

```
Computing GW distance using k= 16 Principal Components
$withinness
    Mild  Control Moderate   Severe 
10.39139 11.53781 10.54993 10.46477 

$betweenness
[1] 11.50008

$gw_dist
[1] 0.01717915
```


```{r eval=FALSE}
computeStructuralMetrics(y_qrilc, group, y[rownames(top.hvp)[1:50],], k = 16)
```

```
Computing GW distance using k= 16 Principal Components
$withinness
    Mild  Control Moderate   Severe 
10.34686 11.84049 10.62378 10.73958 

$betweenness
[1] 11.62664

$gw_dist
[1] 0.008877501
```



`Withinness` tends to be smaller by `msImpute`, which indicates that local structures are better preserved by these two methods. The `gw_dist` over all PCs for the two methods is very similar (rounded to 2 decimals). This suggests the enhancements in `v2-mnar` is just as good as left-censored MNAR methods such as `QRILC`.  Note that `k` is set to the number of samples to capture all dimensions of the data. 


Also note that that, unlike `QRILC`, msImpute `v2-mnar` dose not drastically increase the variance of peptides (measured by squared coefficient of variation) post imputation.
```{r}
par(mfrow=c(2,2))
pcv <- plotCV2(y, main = "data")
pcv <- plotCV2(y_msImpute, main = "msImpute v2-mnar")
pcv <- plotCV2(y_qrilc, main = "qrilc")
```




# SWATH-DIA Case Study: SWATH-MS analysis of Gfi1-mutant bone marrow neutrophils

This study investigates the proteomic alterations in bone marrow neutrophils isolated from 5-8 week old Gfi1+/-, Gfi1K403R/-, Gfi1R412X/-, and Gfi1R412X/R412X mice using the SWATH-MS technique. This dataset consists of 13 DIA (for SWATH) runs on a TripleTOF 5600 plus (SCIEX). Data available from PXD010943. Peak areas extracted from `13DIAs_SWATHprocessing_area_score_FDR_observedRT.xlsx`.^[Accessible via ProteomXchange]

Rows are peptides. Charge state variations are treated as distinct peptide species. Peptides with more than 4 observed intensity values are retained. 

### Normalization

We normalize using `quantile normalization`.
```{r}
data(pxd010943)
y <- pxd010943
# no problematic values for log- transformation
table(is.infinite(data.matrix(log2(y))))

y <- log2(y)
y <- normalizeBetweenArrays(y, method = "quantile")
```

## Determine missing values pattern

```{r}
group <- as.factor(gsub("_[1234]", "", colnames(y)))
group

hdp <- selectFeatures(y, method = "ebm", group = group) 

table(hdp$msImpute_feature)
table(is.nan(hdp$EBM))

table(complete.cases(y))

```
A very small number of peptides (17) tend to be missing in at least one experimental group.

```{r fig.cap="Dropout pattern of informative peptides", fig.align="center"}

# construct matrix M to capture missing entries
M <- ifelse(is.na(y),1,0)
M <- M[hdp$msImpute_feature,]

# plot a heatmap of missingness patterns for the selected peptides



ha_column <- HeatmapAnnotation(group = group)

hm <- Heatmap(M,
column_title = "dropout pattern, columns ordered by dropout similarity",
              name = "Intensity",
              col = c("#8FBC8F", "#FFEFDB"),
              show_row_names = FALSE,
              show_column_names = FALSE,
              cluster_rows = TRUE,
              cluster_columns = TRUE,
              show_column_dend = FALSE,
              show_row_dend = FALSE,
              top_annotation = ha_column,
              row_names_gp =  gpar(fontsize = 7),
              column_names_gp = gpar(fontsize = 8),
              heatmap_legend_param = list(#direction = "horizontal",
              heatmap_legend_side = "bottom",
              labels = c("observed","missing"),
              legend_width = unit(6, "cm")),
         )
hm <- draw(hm, heatmap_legend_side = "left")
```
It can be seen that peptides with structured missing tend to come from the `R412Xhomo` group. Given that a very small number of 
missing peptides exhibit structured missing out of total number of partially observed peptides (17/182), we try both
`method="v2=mnar"` (default, for MNAR data) and `method="v2"` (for MAR data) and compare structural metrics:

## Imputation


```{r}
y_msImpute_mar <- msImpute(y, method = "v2") # no need to specify group if data is MAR.
y_msImpute_mnar <- msImpute(y, method = "v2-mnar", group = group)
```

## Assessment of preservation of local and global structures

In this example, we do not compute `gw_dist` and only rely on `withinness` and `betweenness` metrics to assess imputation.
```{r}

computeStructuralMetrics(y_msImpute_mar, group, y=NULL)
```


```{r}
computeStructuralMetrics(y_msImpute_mnar, group, y = NULL)
```

There do not appear to be substantial difference between the two methods.


Additionally, both of the method preserve variations in the data well:
```{r}
par(mfrow=c(2,2))
pcv <- plotCV2(y, main = "data")
pcv <- plotCV2(y_msImpute_mnar, main = "msImpute v2-mnar")
pcv <- plotCV2(y_msImpute_mar, main = "msImpute v2")
```


# References

Prianichnikov, N., Koch, H., Koch, S., Lubeck, M., Heilig, R., Brehmer, S., Fischer, R., & Cox, J. (2020). MaxQuant Software for Ion Mobility Enhanced Shotgun Proteomics. Molecular & cellular proteomics : MCP, 19(6), 1058–1069. https://doi.org/10.1074/mcp.TIR119.001720

Zhang, X., Deeke, S.A., Ning, Z. et al. Metaproteomics reveals associations between microbiome and intestinal extracellular vesicle proteins in pediatric inflammatory bowel disease. Nat Commun 9, 2873 (2018). https://doi.org/10.1038/s41467-018-05357-4

Muench, D.E., Olsson, A., Ferchen, K. et al. Mouse models of neutropenia reveal progenitor-stage-specific defects. Nature 582, 109–114 (2020). https://doi.org/10.1038/s41586-020-2227-7 

# Session info {-}
```{r echo=FALSE}
sessionInfo()
```
